---
title: "Final_merged_ML"
format: html
editor: visual
---

#Load Libraries
```{r}
#| message: false
#| warning: false

#Loading necessary libraries

# --- Data Import and Basic Handling ---
library(readr)        # For reading CSV and delimited text files
library(janitor)      # For cleaning column names, removing empty rows/columns
library(naniar)       # Handling missing data

# --- Data Wrangling & Cleaning ---
library(dplyr)        # Data manipulation (filter, mutate, group_by, etc.)
library(tidyr)        # Data tidying (pivoting, reshaping)
library(lubridate)    # Date/time manipulation
library(tidyverse)    # Collection of packages for data science


# --- Visualization ---
library(ggplot2)      # Core plotting
library(ggthemes)     # Extra themes for ggplot2
library(ggpubr)       # Publication-ready plots
library(cowplot)      # Combining multiple ggplots
library(corrplot)     # Correlation matrix visualization
library(GGally)       # Pairwise plots (ggpairs)
library (PerformanceAnalytics)
library(plotly)
library(circlize)

# --- Machine Learning Framework ---
library(tidymodels)   # for parsnip, recipes, tune, etc.

# --- Modeling Engines ---
library(xgboost)      # XGBoost implementation
library(ranger)       # Fast Random Forests (optional alternative for comparison)

# --- Feature Engineering ---
library(recipes)      # Included in tidymodels, but useful to load for step-by-step prep
library(textrecipes)  # If you're working with text data

# --- Model Tuning & Validation ---
library(finetune)     # Advanced hyperparameter tuning strategies (racing, ANOVA, etc.)
library(tune)         # For grid/random tuning workflows
library(rsample)      # Data splitting (train/test, cross-validation)

# --- Metrics and Evaluation ---
library(yardstick)    # Evaluation metrics (RMSE, R², MAE, etc.)
library(vip)          # Variable importance plots
library(pROC)         # ROC and AUC curves for classification (if needed)

# --- Parallel Processing ---
library(doParallel)   # Enable parallel computation to speed up tuning/resampling
library(future)

# --- Utilities & Diagnostics ---
library(car)          # Variance inflation factors, etc.
library (skimr)         # For a quick skim/summary of the data
library(DataExplorer) # For EDA

# Load required libraries
library(readr)
library(dplyr)
library(lubridate)
library(ggplot2)
library(tidyr)
library(caret)
```


# Read and merge data (left join)
```{r}

# Read weather data and compute yearly/site means
weather_mean <- read_csv("data/training/weather_data.csv") %>%
  group_by(year, site) %>%
  summarise(
    dayl_s      = mean(dayl_s,      na.rm = TRUE),
    prcp_mm_day = mean(prcp_mm_day, na.rm = TRUE),
    srad_w_m_2  = mean(srad_w_m_2,  na.rm = TRUE),
    swe_kg_m_2   = mean(swe_kg_m_2,   na.rm = TRUE),
    tmax_deg_c  = mean(tmax_deg_c,  na.rm = TRUE),
    tmin_deg_c  = mean(tmin_deg_c,  na.rm = TRUE),
    vp_pa       = mean(vp_pa,       na.rm = TRUE),
    .groups     = "drop"
  )

# 2. Read trait data
trait <- read_csv("data/training/training_trait.csv")

# 3. Join trait with averaged weather
trait_weather <- left_join(trait, weather_mean, by = c("year", "site"))

# 4. Read soil data and clean site codes (remove year suffix)
soil <- read_csv("data/training/training_soil.csv") %>%
  mutate(site = sub(paste0("_", year, "$"), "", site))

# 5. Join trait-weather with soil data
tw_soil <- left_join(trait_weather, soil, by = c("year", "site"))

# 6. Read and clean meta data (drop empty Tmax/Tmin)
meta <- read_csv("data/training/training_meta.csv") %>%
  select(year, site, previous_crop, longitude, latitude)

# 7. Join meta into the combined dataset
tw_sm_meta <- left_join(tw_soil, meta, by = c("year", "site"))

# 8. Remove all rows with any missing values
final_merged <- tw_sm_meta %>%
  drop_na()

# 9. Write out the fully merged and cleaned CSV
#write_csv(final, "merged_meta_trait_weather_soil_clean.csv")

# 10. Preview first rows
#print(head(final))
```

```{r}

# Read weather data and compute yearly/site means
weather_mean = read_csv("data/training/weather_data.csv") %>%
  group_by(year, site) %>%
  summarise(
    dayl_s      = mean(dayl_s,      na.rm = TRUE),
    prcp_mm_day = mean(prcp_mm_day, na.rm = TRUE),
    srad_w_m_2  = mean(srad_w_m_2,  na.rm = TRUE),
    swe_kg_m_2  = mean(swe_kg_m_2,   na.rm = TRUE),
    tmax_deg_c  = mean(tmax_deg_c,  na.rm = TRUE),
    tmin_deg_c  = mean(tmin_deg_c,  na.rm = TRUE),
    vp_pa       = mean(vp_pa,       na.rm = TRUE),
    .groups     = "drop"
  )

# Read trait data
trait = read_csv("data/training/training_trait.csv")

# Join trait with averaged weather data
trait_weather = left_join(trait, weather_mean, by = c("year", "site"))

# Read soil data and clean site codes
trait_weather_soil <- read_csv("data/training/training_soil.csv") %>%
  mutate(site = sub(paste0("_", year, "$"), "", site)) %>%
  left_join(trait_weather, ., by = c("year", "site"))

# 5. Read and clean meta data
meta <- read_csv("data/training/training_meta.csv") %>%
  select(year, site, previous_crop, longitude, latitude)

# 6. Combine all and drop NAs
merged_clean <- trait_weather_soil %>%
  left_join(meta, by = c("year", "site")) %>%
  drop_na() %>%
  select(-site, -replicate, -block, -dayl_s)

# 7. Feature engineering: parse dates and compute Days After Planting (DAP)
merged_clean <- merged_clean %>%
  mutate(
    date_planted   = mdy(date_planted),
    date_harvested = mdy(date_harvested),
    DAP            = as.numeric(date_harvested - date_planted)
  )

merged_clean

# 8. Factorize remaining categorical variables
merged_factored <- merged_clean %>%
  mutate(
    previous_crop = factor(previous_crop),
    hybrid        = factor(hybrid)
  )

# 9. Ensure numeric columns are numeric
decimal_cols <- c(
  "prcp_mm_day","srad_w_m_2","swe_kg_m_2",
  "tmax_deg_c","tmin_deg_c","vp_pa",
  "soilpH","om_pct","soilk_ppm","soilp_ppm",
  "pHom","pk","soilpH^2","om_pct^2","p+k","norm_pk","DAP", "grain_moisture", "yield_mg_ha"
)
merged_factored <- merged_factored %>%
  mutate(across(all_of(decimal_cols), as.numeric))

# 10. Outlier inspection: boxplots for numeric predictors
num_vars <- merged_factored %>% select(all_of(decimal_cols))
boxplot_data <- num_vars %>%
  pivot_longer(cols = everything(), names_to = "variable", values_to = "value")

ggplot(boxplot_data, aes(x = variable, y = value)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Boxplots for Numeric Predictors")

```

```{r predictors visualization}

# 11. Correlation matrix among continuous predictors
corr_mat <- cor(num_vars, use = "pairwise.complete.obs")
corr_mat


#correlation matrix 
corrplot(
  corr_mat,
  method    = "color",     # "circle", "number", "shade", etc.
  type      = "upper",     # show only upper triangle
  order     = "hclust",    # cluster variables
  tl.col    = "black",     # label color
  tl.srt    = 45,          # label rotation
  addCoef.col = "white",   # add correlation coefficients
  diag      = FALSE        # hide diagonal
)


# combined scatter + histogram + corr matrix view
chart.Correlation(
  num_vars,
  histogram    = TRUE,
  pch          = 19,
  method       = "pearson"  # or "spearman"
)


# Chord Diagram of Strong Correlations 
strong = ifelse(abs(corr_mat) > 0.7, corr_mat, 0)
chordDiagram(strong, transparency = 0.5)


#Interactive 3D correlation Surface
plot_ly(
  z = corr_mat,
  x = colnames(corr_mat),
  y = colnames(corr_mat),
  type = "surface"
)

```


#XGBoost ML workflow 
## 1. Pre-processing
```{r }

# Set seed for reproducibility
set.seed(1996)

# Create 70/30 train-test split, stratified by yield
ml_split <- initial_split(
  data   = num_vars,
  prop   = 0.7,
  strata = yield_mg_ha # Stratify by target variable
)

# Review split
ml_split


```
### a. Data split
```{r train data}
train_data = training(ml_split)  # 70% of data
train_data 
```


```{r test data}

train_data = testing(ml_split)  # 30% of data
train_data

```

### b. Distribution of target variable
```{r density plot of target variable}
ggplot() +
  geom_density(data = train_data, 
               aes(x = yield_mg_ha),
               color = "red") +
  geom_density(data = test_data, 
               aes(x = yield_mg_ha),
               color = "blue") # it shows approxmately normal distribution
  
```

### c. Data processing with recipe

```{r}
library(recipes)

ml_recipe <- recipe(yield_mg_ha ~ ., data = train_data) %>%
  # Center & scale all remaining predictors
  step_normalize(all_predictors()) %>%
  # Apply PCA to those normalized predictors
  # Option A: keep PCs explaining 95% of variance
  step_pca(all_predictors(), threshold = 0.95)
  # Option B: keep a fixed number of components, e.g. 5
  # step_pca(all_predictors(), num_comp = 5)

ml_recipe

```
```{r}
# Prep the recipe to estimate any required statistics
ml_prep <- ml_recipe %>% 
  prep()

# Examine preprocessing steps
ml_prep
```

## 2. Training
### a. Model specification

```{r XGBoost specs}
# XGBoost regression model spec with hyperparameter tuning
xgb_spec <- boost_tree(
  trees      = tune(),  # Total number of boosting iterations
  tree_depth = tune(),  # Maximum depth of each tree
  min_n      = tune(),  # Minimum number of samples required to split a node
  learn_rate = tune()   # Step-size shrinkage (learning rate)
) %>%
  set_engine("xgboost") %>%  # Specify the XGBoost backend
  set_mode("regression")     # Fit a regression model

# Print the specification
xgb_spec

```

### b. Cross-validation setup

```{r 10-fold CV}
set.seed(1996)

resampling_foldcv = vfold_cv(train_data, 
                             v = 10) #10-fold cross validation

resampling_foldcv
resampling_foldcv$splits[[1]]
```

### c. Hyperparameter grid with Latin Hypercube Sampling
```{r XGBoost grid}
xgb_grid = grid_latin_hypercube(
  tree_depth(),
  min_n(),
  learn_rate(),
  trees(),
  size = 100
)

xgb_grid
```

```{r XGB grid visualization}
ggplot(data = xgb_grid,
       aes(x = tree_depth, 
           y = min_n)) +
  geom_point(aes(color = factor(learn_rate),
                 size = trees),
             alpha = .5,
             show.legend = FALSE)
```
## 3. Model Tuning

```{r xgb_grid_result}
#for reproducibility
set.seed(1996)

# Parallel processing
registerDoParallel(cores = parallel::detectCores() - 1)

# Perform grid search tuning
xgb_res <- tune_grid(
  object = xgb_spec,
  preprocessor = ml_recipe, #providing recipe here
  resamples = resampling_foldcv, # providing re-sampling CV
  grid = xgb_grid, # providing grid specs
  control = control_race(save_pred = TRUE)
)

# Stop parallel processing
stopImplicitCluster()

# Play a sound when done
beepr::beep()

# Access metrics
xgb_res$.metrics[[2]]
```

## 4. Select Best Models
```{r}
# Based on lowest RMSE
best_rmse <- xgb_res %>% 
  select_best(metric = "rmse")%>% 
  mutate(source = "best_rmse")

best_rmse
```

```{r}
# Based on lowers RMSE within 1% loss
best_rmse_pct_loss <- xgb_res %>% 
  select_by_pct_loss("min_n",
                     metric = "rmse",
                     limit = 1
                     )%>% 
  mutate(source = "best_rmse_pct_loss")

best_rmse_pct_loss
```
```{r}
# Based on lowest RMSE within 1 se
best_rmse_one_std_err <- xgb_res %>% 
  select_by_one_std_err(metric = "rmse",
                        eval_time = 100,
                        trees
                        )%>% 
  mutate(source = "best_rmse_one_std_err")

best_rmse_one_std_err
```
```{r}
# Based on greatest R2
best_r2 <- xgb_res %>% 
  select_best(metric = "rsq")%>% 
  mutate(source = "best_r2")

best_r2
```

```{r}
# Based on lowers R2 within 1% loss
best_r2_pct_loss <- xgb_res %>% 
  select_by_pct_loss("min_n",
                     metric = "rsq",
                     limit = 1
                     ) %>% 
  mutate(source = "best_r2_pct_loss")

best_r2_pct_loss
```
```{r}
# Based on lowest R2 within 1 se
best_r2_one_std_error <- xgb_res %>% 
  select_by_one_std_err(metric = "rsq",
                        eval_time = 100,
                        trees
                        ) %>%
  mutate(source = "best_r2_one_std_error")

best_r2_one_std_error
```
## Compare and Finalize Model
```{r}
best_rmse %>% 
  bind_rows(best_rmse_pct_loss, 
            best_rmse_one_std_err, 
            best_r2, 
            best_r2_pct_loss, 
            best_r2_one_std_error)
```

## 5. Final Specification

```{r}
final_spec <- boost_tree(
  trees = best_r2$trees,           # Number of boosting rounds (trees)
  tree_depth = best_r2$tree_depth, # Maximum depth of each tree
  min_n = best_r2$min_n,           # Minimum number of samples to split a node
  learn_rate = best_r2$learn_rate  # Learning rate (step size shrinkage)
) %>%
  set_engine("xgboost") %>%
  set_mode("regression")

final_spec
```

## 6. Final Fit and Predictions
## Validation

```{r final_fit}
set.seed(1996)
final_fit <- last_fit(final_spec,
                ml_recipe,
                split = ml_split)

final_fit %>%
  collect_predictions()
```
## 7. Evaluate on Test Set
```{r}
final_fit %>%
  collect_metrics()
```

## 8. Evaluate on Training Set
```{r}
final_spec %>%
  fit(yield_mg_ha ~ .,
      data = bake(ml_prep, 
                  train_data)) %>%
  augment(new_data = bake(ml_prep, 
                          train_data)) %>% 
  rmse(yield_mg_ha, .pred) %>%
  bind_rows(
    
    
# R2
final_spec %>%
  fit(yield_mg_ha ~ .,
      data = bake(ml_prep, 
                  train_data)) %>%
  augment(new_data = bake(ml_prep, 
                          train_data)) %>% 
  rsq(yield_mg_ha, .pred))
```

## 9. Predicted vs Observed Plot
```{r}
preds_in_range <- 
  final_fit %>%
  collect_predictions() %>%
  filter(
    between(yield_mg_ha, 20, 40),
    between(.pred,        20, 40),
    !is.na(.pred)
  )

ggplot(preds_in_range, aes(x = yield_mg_ha, y = .pred)) +
  geom_point() +
  geom_abline() +
  geom_smooth(method = "lm", formula = y ~ x) +
  # now limits aren’t needed, since everything is already in [20,40]:
  labs(
    x = "Observed Yield (Mg ha⁻¹)",
    y = "Predicted Yield (Mg ha⁻¹)",
    title = "Observed vs. Predicted Cotton Yield"
  )

```
## 10. Variable Importance
```{r}
final_spec %>%
  fit(yield_mg_ha ~ .,
         data = bake(ml_prep, train_data)) %>% 
    vi() %>%
  mutate(
    Variable = fct_reorder(Variable, 
                           Importance)
  ) %>%
  ggplot(aes(x = Importance, 
             y = Variable)) +
  geom_col() +
  scale_x_continuous(expand = c(0, 0)) +
  labs(y = NULL)
```


#Summary




